---
title: "The Total Decedents of Shelter Residents in Toronto by Year"
subtitle: "The Decedents significantly increased during the pandemic"
author: 
  - Haobo Ren
thanks: "Code and data are available at: https://github.com/HaoboRrrr/The_total_decedents_of_shelter_residents_in_Toronto"
date: 23-09-2024
date-format: long
abstract: "Study of the number of decedents of shelter residents in Toronto by year."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
```


# Introduction
Most people don't have a clear picture of what life is like for the homeless and those living in shelters. @krajewska2016differing It is often known that the homeless population has higher incidence of sickness and early mortality.The City is dedicated to recognizing new and developing problems in the shelter system and acting quickly to address them. @death_of_shelter_residents This paper aims to analyse 
the trend and level of deaths numbers in shelter in a yearly basis from 2007 to 2024. \newline

This paper uses R @citeR to analyse involved data. In the process of analyzing and visualizing the data set, the flowing R packages are used: tidyverse @tidyverse, janitor @janitor, ggplot2 @ggplot2, opendatatoronto @opendatatoronto and knitr @knitr\newline

In the following Data section will exhibit the years and its corresponding total decedents in the shelters in Toronto area, and how the visulized data looks like. \newline

# Data {#sec-data}
The Toronto Shelter and Support Services Division has collected this information since 2007 and conducts an annual review of the data. (@fig-data)

```{r}
#| label: fig-data
#| fig-cap: Data of total decedents each year
#| echo: false
data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
knitr::kable(data)
```

In this data, we can see that there were many decedents in the shelter each year from 2007 to 2024ytd. This means that people living in shelters are at greater risk than the general public.They may experience the threat of drug abuse, violence, illegal guns. @brown2011spread This raises a new question: Why does the government condone people living in shelters and even provide them with drugs? Isn't this making their lives worse? \newline

After viewing the actual data, visualizing the data can provide us with a different perspective. (@fig-graph)
```{r}
#| label: fig-graph
#| fig-cap: Barplot of Total decedents each year
#| fig-width: 10
#| echo: false
data |> 
  ggplot(aes(x = year, y = total_decedents)) +
  geom_bar(stat="identity") +
  ggtitle("Bar plot of shelter residents' total decedents from 2007 to 2024 by year") +
  xlab("Years") +
  ylab("Total decedents")

```
The average deaths per year prior to 2020 was about 25, this number spikes during 2020 to 2023. During the pandemic. The operation of the shelter has been greatly affected during the epidemic. Toronto's medical resources are so stretched that they are unable to care for the people in the shelter. @khandor_2011_access


# Conclusion
From these data, we know that people living in shelters are always living in danger, their life expectancy is shorter than most people in society. They are also the most likely to be forgotten when society is hit. Perhaps what the government needs to do is to guide such people to reintegrate into society and get back to work, rather than indulge them and think that they have fulfilled their obligations by just providing food.






\newpage


# References


